---
title: "rep_data-PA02: The Most Damaging Types Of Weather Events in the USA"
author: "Ronald Stalder"
output: 
  html_document:
    keep_md: true
---

## &nbsp;
## Synopsis

This report was produced by the author as a student's exercise during the online course ["Data Science - Reproducible Research"](https://www.coursera.org/course/repdata) taken in Dezember 2014.

Analysing the data from the **U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database** for the years 1950 to 2011, the following **event types** were found most damaging regarding:  
* **human fatalities and injuries:** Tornadoes are, by far, the worst.    
* **material damage to properties and crops:** Hurricanes, Flash Floods and Tornadoes are the leaders    
&nbsp;  
The database is a mess and needs a lot of consideration regarding **cleaning the data**. 
Especially, there are many non standard **Event Types** and at least one event with largly exagerated damgae values.


## &nbsp;
## Data Processing

#### initialize

```{r setup, results='hide', error=FALSE, message=FALSE}

StormDataFile <- "StormData.csv.bz2"

library(data.table)
library(plyr)

```

#### &nbsp;
#### download the data file

* **note:** run this once, then set chunk option `eval=FALSE`

```{r download-1, eval=FALSE}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,StormDataFile,method="curl")

```
```{r download-2}

print(paste("the file was downloaded on:", file.info(StormDataFile)$mtime))

```

* **note:**  
    This file had been prepared and uploaded by the course staff. It contains events from
    1950 to 2011. For more recent data, and for information about the history of the data,
    see the [NOOA - Storm Events Database](http://www.ncdc.noaa.gov/stormevents/details.jsp?type=collection).

#### &nbsp;
#### read the data file into the data table STORM_RAW

```{r readcsv, cache=TRUE}

STORM_RAW <- as.data.table(
                 read.csv(StormDataFile,
                          stringsAsFactors=FALSE)
         )

# str(STORM_RAW)
head(STORM_RAW, 1)

```

#### &nbsp;
#### cleaning the data

1. We are only interested in events that caused damage (human or material). Remove all 
entries without any damage reported. Also, remove all columns not needed for the analysis.

```{r clean-1, cache=TRUE, results='hold'}

# remove rows without damage reported
STORM <- STORM_RAW[!(FATALITIES==0 & INJURIES==0 & PROPDMG==0 & CROPDMG==0), ]
print(paste("this removed", dim(STORM_RAW)[1]-dim(STORM)[1], "entries, leaving", dim(STORM)[1]))
print("")

# keep only needed columns
STORM <- STORM[, list(BGN_DATE, EVTYPE, FATALITIES, INJURIES,
                      PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REFNUM)]
# for the begin-date, only the year is needed
STORM <- STORM[, EventYear:=sub(".*/([0-9]{4}).*", "\\1", BGN_DATE)][, BGN_DATE:=NULL]

str(STORM)

```

&nbsp;  
2. Tidy up the prop and crop damage values, putting them into one numeric column each

```{r clean-2a, cache=TRUE}

entries <- dim(STORM)[1]

# permitted exponent factors are "",K,M,B but allow also for lowercase
permittedEXP <- list("","K","M","B")
STORM <- STORM[lapply(PROPDMGEXP, toupper) %in% permittedEXP, ]
STORM <- STORM[lapply(CROPDMGEXP, toupper) %in% permittedEXP, ]

# insert the numeric columns
STORM <- STORM[ ,`:=`(PropDamage=as.numeric(PROPDMG),
                      CropDamage=as.numeric(CROPDMG))]

# multiply by EXP
STORM <- STORM[toupper(PROPDMGEXP)=="K", PropDamage:=1000*PropDamage]
STORM <- STORM[toupper(PROPDMGEXP)=="M", PropDamage:=1000000*PropDamage]
STORM <- STORM[toupper(PROPDMGEXP)=="B", PropDamage:=1000000000*PropDamage]

STORM <- STORM[toupper(CROPDMGEXP)=="K", CropDamage:=1000*CropDamage]
STORM <- STORM[toupper(CROPDMGEXP)=="M", CropDamage:=1000000*CropDamage]
STORM <- STORM[toupper(CROPDMGEXP)=="B", CropDamage:=1000000000*CropDamage]

print(paste("this removed", entries-dim(STORM)[1], "entries, leaving", dim(STORM)[1]))

summary(STORM[, list(PropDamage, CropDamage)])

```

* **note:**  
    There are only a few non standard exponents and, for the purpose of this assignment,
    I decided to **ignore them** (remove them from the dataset).  
    
&nbsp;  
**looking for outliers:**

```{r clean-2b, cache=TRUE, results='hold'}

summary(STORM[, list(FATALITIES, INJURIES)])
print(" ")
summary(STORM[, list(PropDamage, CropDamage)])

```

The maximum values for Prop and Crop damages look rather suspicious ($ 115 **billions??**). Let's find out where these come from:

```{r clean-2c, cache=TRUE}

# get the REFNUM of the 2 max for prop and crop
pRef <- STORM[which.max(PropDamage), REFNUM]
cRef <- STORM[which.max(CropDamage), REFNUM]

# print the complete event reports
print(STORM_RAW[REFNUM==pRef], )
# print(STORM_RAW[REFNUM==cRef], )

```
  
* Well, from the remarks of the first event (115bn PropDamage), follows clearly, that this should be **millions**, not billions.  
* As for the second, the 5bn CropDamage, this could be plausible as thousands of acres stayed under water (I did not print it out as the description is very long).  
* Correct the first one:

```{r clean-2d, cache=TRUE}

# millions not billions:
STORM <- STORM[REFNUM==pRef, `:=`(PropDamage=PropDamage/1000,CropDamage=CropDamage/1000)]

```

&nbsp;  
3. **Storm Data Event Types:** As per [documentation, table 2.1.1 page 6](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), the following event types are permitted  (*manually* copy/paste/edited from the .pdf):

```{r clean-3a, cache=TRUE}

permitted_EVTYPES <- list(
        "Astronomical Low Tide",
        "Avalanche",
        "Blizzard",
        "Coastal Flood",
        "Cold/Wind Chill",
        "Debris Flow",
        "Dense Fog",
        "Dense Smoke",
        "Drought",
        "Dust Devil",
        "Dust Storm",
        "Excessive Heat",
        "Extreme Cold/Wind Chill",
        "Flash Flood",
        "Flood",
        "Frost/Freeze",
        "Funnel Cloud",
        "Freezing Fog",
        "Hail",
        "Heat",
        "Heavy Rain",
        "Heavy Snow",
        "High Surf",
        "High Wind",
        "Hurricane (Typhoon)",
        "Ice Storm",
        "Lake-Effect Snow",                 
        "Lakeshore Flood",
        "Lightning",
        "Marine Hail",
        "Marine High Wind",
        "Marine Strong Wind",                                        
        "Marine Thunderstorm Wind",
        "Rip Current",
        "Seiche",
        "Sleet",
        "Storm Surge/Tide",
        "Strong Wind",
        "Thunderstorm Wind",
        "Tornado",
        "Tropical Depression",
        "Tropical Storm",
        "Tsunami",
        "Volcanic Ash",
        "Waterspout",
        "Wildfire",
        "Winter Storm",
        "Winter Weather" 
  )

# convert all event types to upper case
STORM <- STORM[ ,EVTYPE:=toupper(EVTYPE)]
permitted_EVTYPES <- lapply(permitted_EVTYPES, toupper)

# estimate what we are missing with the non standard EVTYPE
number_illegal <- dim(STORM[!EVTYPE %in% permitted_EVTYPES,])[1]
ALL <- STORM[, `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
NON <- STORM[!EVTYPE %in% permitted_EVTYPES, ][,
             `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
pctNb <- round(100*number_illegal/dim(STORM)[1], 2)                 # %entries
pctHD <- round(100*(NON[1,sF]+NON[1,sI])/(ALL[1,sF]+ALL[1,sI]), 2)  # %Human damage
pctMD <- round(100*(NON[1,sP]+NON[1,sC])/(ALL[1,sP]+ALL[1,sC]), 2)  # %Material damage

```

There are `r number_illegal` entries with non permitted event types. They represent:

* `r pctNb`% of the number of entries
* `r pctHD`% of the total of fatalities + injuries
* `r pctMD`% of the total of material damage (prop + crop)
    
So let's take a closer look at them:
    
```{r clean-3b}

# list the non permitted events
sort(unique(STORM[!EVTYPE %in% permitted_EVTYPES, EVTYPE]))

```

Correct obvious errors:
    
```{r clean-3c}

# remove leading spaces
STORM <- STORM[ ,EVTYPE:=sub("^ *([^ ].*)", "\\1", EVTYPE)]
# mispelling of THUNDERSTORM and abbrev. TSTM
STORM <- STORM[, EVTYPE:=sub("THUDERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDEERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERESTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERSTROM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERSTORMS", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("TUNDERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("TUNDERTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("TSTM", "THUNDERSTORM", EVTYPE)]
# mispelled TORNADO
STORM <- STORM[, EVTYPE:=sub("TORNDAO", "TORNADO", EVTYPE)]

# some obvious adaptions: remove storm names, clean up floods, etc
STORM <- STORM[, EVTYPE:=sub("^HIGH *WIND.*", "HIGH WIND", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^HURRICANE.*", "HURRICANE (TYPHOON)", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^TYPHOON.*", "HURRICANE (TYPHOON)", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^THUNDERSTORM.*", "THUNDERSTORM WIND", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^SEVERE THUNDERSTORM.*", "THUNDERSTORM WIND", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^TORNADO.*", "TORNADO", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^TROPICAL STORM.*", "TROPICAL STORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FLOOD.*FLASH.*", "FLASH FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FLASH FLOOD.*", "FLASH FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^RIVER.*FLOOD.*", "FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^URBAN FLOOD.*", "FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FLOOD.*", "FLASH FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^RIP CURRENT.*", "RIP CURRENT", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^HAIL.*", "HAIL", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("SMALL HAIL", "HAIL", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^WILD.*", "WILDFIRE", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^STORM SURGE$", "STORM SURGE/TIDE", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^EXTREME COLD$", "EXTREME COLD/WIND CHILL", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FOG$", "DENSE FOG", EVTYPE)]

# estimate what we are missing now
number_illegal <- dim(STORM[!EVTYPE %in% permitted_EVTYPES,])[1]
ALL <- STORM[, `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
NON <- STORM[!EVTYPE %in% permitted_EVTYPES, ][,
             `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
pctNb <- round(100*number_illegal/dim(STORM)[1], 2)                 # %entries
pctHD <- round(100*(NON[1,sF]+NON[1,sI])/(ALL[1,sF]+ALL[1,sI]), 2)  # %Human damage
pctMD <- round(100*(NON[1,sP]+NON[1,sC])/(ALL[1,sP]+ALL[1,sC]), 2)  # %Material damage

# remove the non-permitted EVTYPE rows
STORM <- STORM[EVTYPE %in% permitted_EVTYPES, ]

```

There remained `r number_illegal` entries with non permitted event types, representing:

* `r pctNb`% of the number of entries
* `r pctHD`% of the total of fatalities + injuries
* `r pctMD`% of the total of material damage (prop + crop)

That's good enough, so they have been removed from the dataset. Let's begin the analysis.

    

## &nbsp;
## Results

### &nbsp;
### Introduction

From the [NOOA - Storm Events Database - EventTypes](http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype) follows that the 48 defined Event Types are only registered since 1996, prior to this only a small subset of event types were recorded. This will affect the list (see tornadoes below) but, for the purpose here, I will leave it at this.  

### &nbsp;
### Human Damage: Fatalities and Injuries

The problem here is to find the most damaging event types for fatalities and injuries *together*. Just adding them up doesn't make sense, neither ethically nor economically. First, let's find out whether the problem really arises:

```{r human-1, results='hold'}

# get total fatalities and injuries
HUMD <- unique(STORM[ ,`:=`(sF=sum(FATALITIES),sI=sum(INJURIES)), 
                       by=EVTYPE][, list(EVTYPE, sF, sI)],
                by="EVTYPE")

print("10 worst event types regarding fatalities:")
setorder(HUMD,-sF)
HUMD[1:10, ]

print("")
print("10 worst event types regarding injuries:")
setorder(HUMD,-sI)
HUMD[1:10, ]

```

So, yes, the problem is there. What is worse  
- avalanche (224 killed, 170 injured), or  
- wildfire (90 killed, 1606 injured)  ??  

It would go far beyond the scope of this exercise to resolve this question. Despite the
problematic of doing so, I will add the two together to classify by the
*total number of people affected*.  

This is the list and the corresponding bar plot:  

```{r human-2}

# get total fatalities and injuries and total people affected
HUMD <- unique(STORM[ ,`:=`(sF= sum(FATALITIES), sI=sum(INJURIES), totAff=sF+sI), by=EVTYPE]
                    [, list(EVTYPE, sF, sI, totAff)],
                by="EVTYPE")
# limit to 10 worst
setorder(HUMD,-totAff)
HUMD <- HUMD[1:10, ]
HUMD

# bar plot
hdm <- as.matrix(HUMD[ ,list(sF,sI)])
colnames(hdm) <- c("Fatalities","Injuries")
rownames(hdm) <- as.vector(HUMD[ ,EVTYPE])
hdm <- t(hdm)  # rows <> cols
barplot(hdm, 
        col = c("navyblue", "royalblue2"), 
        ylim = c(0, 10000),
        beside = FALSE, las = 3, 
        cex.axis = 0.7, cex.names = 0.4,
        main = "Most damaging event types:\nHuman Damage 1950-2011\n(worst 10)", 
        ylab = "Total number of affected people", 
        legend.text = rownames(hdm))

```

* **note:**  
    *"Tornado"* is the most damaging event type, by far. It affects nearly 10 times more
    people than the next most damaging event type, *"Thunderstorm Wind"*. I deliberatly set
    the y-axis to a max of 10,000 and let the tornadoes run out of the picture - otherwise
    there would be a poor plot and I didn't want to use a log() scale.



### &nbsp;
### Material Damage: Losses in Properties and Crops

This is the list and the corresponding bar plot:  

```{r material-1}

# get total fatalities and injuries and total people affected
MATD <- unique(STORM[ ,`:=`(Prop=sum(PropDamage), Crop=sum(CropDamage)), 
                            by=EVTYPE]
                    [, list(EVTYPE, Prop, Crop)],
                by="EVTYPE")
MATD <- MATD[ ,Total:=Prop+Crop]
# limit to 10 worst
setorder(MATD,-Total)
MATD <- MATD[1:10, ]
# reduce to millions
MATD <- MATD[ ,`:=`(Prop=Prop/1000000, Crop=Crop/1000000, Total=Total/1000000)] 
MATD

# bar plot
mdm <- as.matrix(MATD[ ,list(Prop,Crop)])
rownames(mdm) <- as.vector(MATD[ ,EVTYPE])
mdm <- t(mdm)  # rows <> cols
barplot(mdm, 
        col = c("navyblue", "royalblue2"), 
        # ylim = c(0, 10000),
        beside = FALSE, las = 3, 
        cex.axis = 0.7, cex.names = 0.4,
        main = "Most damaging event types:\nMaterial Damage 1950-2011\n(worst 10, in million US$)", 
        ylab = "Total Damage in Million US$", 
        legend.text = rownames(mdm))

```


## &nbsp;
## System and R-Packages Information

```{r sys-1, results='hold'}

if (Sys.which("uname") != "") system("uname -srpi", intern=TRUE)
if (Sys.which("lsb_release") != "") 
    print(paste("Ubuntu:",system("lsb_release -rc", intern=TRUE)[1]))
#print(paste("Rstudio version:", rstudio::versionInfo()$version))  # does not work in Rmd

```
```{r sys-2}

sessionInfo()

```

The RMarkdown file has been published as a [Github Project](https://github.com/ronasta/RepData_PeerAssessment2)  
  
The report was produced using `RStudio/knittr`  
**on `r format(Sys.time(), "%F at %H:%M:%S (%Z, GMT%z)")`**  
