---
title: "rep_data-PA02: The Most Damaging Types Of Weather Events in the USA"
author: "Ronald Stalder"
output: 
  html_document:
    keep_md: true
---

## &nbsp;
## Synopsis

This report was produced by the author as a student's exercise during the online course ["Data Science - Reproducible Research"](https://www.coursera.org/course/repdata) taken in Dezember 2014.

Analysing the data from the **U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database** for the years 1950 to 2011, the following **event types** were found most damaging regarding:  
* **human fatalities and injuries:**    
* **material damage to properties and crops:**    
&nbsp;  
The database is a mess and needs a lot of consideration regarding **cleaning the data**. 
Especially, there are many non standard **Event Types**.


## &nbsp;
## Data Processing

#### initialize

```{r setup, results='hide'}

StormDataFile <- "StormData.csv.bz2"

library(data.table)

```

#### &nbsp;
#### download the data file

* **note:** run this once, then set chunk option `eval=FALSE`

```{r download-1, eval=FALSE}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,StormDataFile,method="curl")

```
```{r download-2}

print(paste("the file was downloaded on:", file.info(StormDataFile)$mtime))

```

* **note:**  
    This file had been prepared and uploaded by the course staff. It contains events from
    1950 to 2011. For more recent data, and for information about the history of the data,
    see the [NOOA - Storm Events Database](http://www.ncdc.noaa.gov/stormevents/details.jsp?type=collection).

#### &nbsp;
#### read the data file into the data table STORM_RAW

```{r readcsv, cache=TRUE}

STORM_RAW <- as.data.table(
                 read.csv(StormDataFile,
                          stringsAsFactors=FALSE)
         )

# str(STORM_RAW)
head(STORM_RAW, 1)

```

#### &nbsp;
#### cleaning the data

1. We are only interested in events that caused damage (human or material). Remove all 
entries without any damage reported. Also, remove all columns not needed for the analysis.

```{r clean-1, cache=TRUE, results='hold'}

# remove rows without damage reported
STORM <- STORM_RAW[!(FATALITIES==0 & INJURIES==0 & PROPDMG==0 & CROPDMG==0), ]
print(paste("this removed", dim(STORM_RAW)[1]-dim(STORM)[1], "entries, leaving", dim(STORM)[1]))
print("")

# keep only needed columns
STORM <- STORM[, list(BGN_DATE, EVTYPE, FATALITIES, INJURIES,
                      PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REFNUM)]
# for the begin-date, only the year is needed
STORM <- STORM[, Year:=sub(".*/([0-9]{4}).*", "\\1", BGN_DATE)][, BGN_DATE:=NULL]

str(STORM)

```

&nbsp;  
2. Tidy up the prop and crop damage values, putting them into one numeric column each

```{r clean-2, cache=TRUE}

entries <- dim(STORM)[1]

# permitted exponent factors are "",K,M,B but allow also for lowercase
permittedEXP <- list("","K","M","B")
STORM <- STORM[lapply(PROPDMGEXP, toupper) %in% permittedEXP, ]
STORM <- STORM[lapply(CROPDMGEXP, toupper) %in% permittedEXP, ]

# insert the numeric columns
STORM <- STORM[ ,`:=`(PropDamage=as.numeric(PROPDMG),
                      CropDamage=as.numeric(CROPDMG))]

# multiply by EXP
STORM <- STORM[toupper(PROPDMGEXP)=="K", PropDamage:=1000*PropDamage]
STORM <- STORM[toupper(PROPDMGEXP)=="M", PropDamage:=1000000*PropDamage]
STORM <- STORM[toupper(PROPDMGEXP)=="B", PropDamage:=1000000000*PropDamage]

STORM <- STORM[toupper(CROPDMGEXP)=="K", CropDamage:=1000*CropDamage]
STORM <- STORM[toupper(CROPDMGEXP)=="M", CropDamage:=1000000*CropDamage]
STORM <- STORM[toupper(CROPDMGEXP)=="B", CropDamage:=1000000000*CropDamage]

print(paste("this removed", entries-dim(STORM)[1], "entries, leaving", dim(STORM)[1]))

summary(STORM[, list(PropDamage, CropDamage)])

```

3. **Storm Data Event Types:** As per [documentation, table 2.1.1 page 6](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), the following event types are permitted  (copy/paste/edited from the .pdf):

```{r clean-3a, cache=TRUE}

permitted_EVTYPES <- list(
        "Astronomical Low Tide",
        "Avalanche",
        "Blizzard",
        "Coastal Flood",
        "Cold/Wind Chill",
        "Debris Flow",
        "Dense Fog",
        "Dense Smoke",
        "Drought",
        "Dust Devil",
        "Dust Storm",
        "Excessive Heat",
        "Extreme Cold/Wind Chill",
        "Flash Flood",
        "Flood",
        "Frost/Freeze",
        "Funnel Cloud",
        "Freezing Fog",
        "Hail",
        "Heat",
        "Heavy Rain",
        "Heavy Snow",
        "High Surf",
        "High Wind",
        "Hurricane (Typhoon)",
        "Ice Storm",
        "Lake-Effect Snow",                 
        "Lakeshore Flood",
        "Lightning",
        "Marine Hail",
        "Marine High Wind",
        "Marine Strong Wind",                                        
        "Marine Thunderstorm Wind",
        "Rip Current",
        "Seiche",
        "Sleet",
        "Storm Surge/Tide",
        "Strong Wind",
        "Thunderstorm Wind",
        "Tornado",
        "Tropical Depression",
        "Tropical Storm",
        "Tsunami",
        "Volcanic Ash",
        "Waterspout",
        "Wildfire",
        "Winter Storm",
        "Winter Weather" 
  )

# convert all event types to upper case
STORM <- STORM[ ,EVTYPE:=toupper(EVTYPE)]
permitted_EVTYPES <- lapply(permitted_EVTYPES, toupper)

# estimate what we are missing with the non standard EVTYPE
number_illegal <- dim(STORM[!EVTYPE %in% permitted_EVTYPES,])[1]
ALL <- STORM[, `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
NON <- STORM[!EVTYPE %in% permitted_EVTYPES, ][,
             `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
pctNb <- round(100*number_illegal/dim(STORM)[1], 2)                 # %entries
pctHD <- round(100*(NON[1,sF]+NON[1,sI])/(ALL[1,sF]+ALL[1,sI]), 2)  # %Human damage
pctMD <- round(100*(NON[1,sP]+NON[1,sC])/(ALL[1,sP]+ALL[1,sC]), 2)  # %Material damage

```

There are `r number_illegal` entries with non permitted event types. They represent:

* `r pctNb`% of the number of entries
* `r pctHD`% of the total of fatalities + injuries
* `r pctMD`% of the total of material damage (prop + crop)
    
So let's take a closer look at them:
    
```{r clean-3b}

# list the non permitted events
sort(unique(STORM[!EVTYPE %in% permitted_EVTYPES, EVTYPE]))

```

Correct obvious errors:
    
```{r clean-3c}

# remove leading spaces
STORM <- STORM[ ,EVTYPE:=sub("^ *([^ ].*)", "\\1", EVTYPE)]
# mispelling of THUNDERSTORM and abbrev. TSTM
STORM <- STORM[, EVTYPE:=sub("THUDERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDEERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERESTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERSTROM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERSTORMS", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("TUNDERSTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("TUNDERTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("THUNDERTORM", "THUNDERSTORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("TSTM", "THUNDERSTORM", EVTYPE)]
# mispelled TORNADO
STORM <- STORM[, EVTYPE:=sub("TORNDAO", "TORNADO", EVTYPE)]

# some obvious adaptions: remove storm names, clean up floods, etc
STORM <- STORM[, EVTYPE:=sub("^HIGH *WIND.*", "HIGH WIND", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^HURRICANE.*", "HURRICANE (TYPHOON)", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^TYPHOON.*", "HURRICANE (TYPHOON)", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^THUNDERSTORM.*", "THUNDERSTORM WIND", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^SEVERE THUNDERSTORM.*", "THUNDERSTORM WIND", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^TORNADO.*", "TORNADO", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^TROPICAL STORM.*", "TROPICAL STORM", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FLOOD.*FLASH.*", "FLASH FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FLASH FLOOD.*", "FLASH FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^RIVER.*FLOOD.*", "FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^URBAN FLOOD.*", "FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^FLOOD.*", "FLASH FLOOD", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^HAIL.*", "HAIL", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("SMALL HAIL", "HAIL", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^WILD.*", "WILDFIRE", EVTYPE)]
STORM <- STORM[, EVTYPE:=sub("^STORM SURGE$", "STORM SURGE/TIDE", EVTYPE)]

# estimate what we are missing now
number_illegal <- dim(STORM[!EVTYPE %in% permitted_EVTYPES,])[1]
ALL <- STORM[, `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
NON <- STORM[!EVTYPE %in% permitted_EVTYPES, ][,
             `:=`(sF=sum(FATALITIES),sI=sum(INJURIES),sP=sum(PropDamage),sC=sum(CropDamage))
             ][1, list(sF,sI,sP,sC)]
pctNb <- round(100*number_illegal/dim(STORM)[1], 2)                 # %entries
pctHD <- round(100*(NON[1,sF]+NON[1,sI])/(ALL[1,sF]+ALL[1,sI]), 2)  # %Human damage
pctMD <- round(100*(NON[1,sP]+NON[1,sC])/(ALL[1,sP]+ALL[1,sC]), 2)  # %Material damage

# remove the non-permitted EVTYPE rows
STORM <- STORM[EVTYPE %in% permitted_EVTYPES, ]

```

There remained `r number_illegal` entries with non permitted event types, representing:

* `r pctNb`% of the number of entries
* `r pctHD`% of the total of fatalities + injuries
* `r pctMD`% of the total of material damage (prop + crop)

That's good enough, so they have been removed from the dataset. Let's begin the analysis.

    

## &nbsp;
## Results


## &nbsp;
## System and R-Packages Information

```{r sys-1, results='hold'}

if (Sys.which("uname") != "") system("uname -srpi", intern=TRUE)
if (Sys.which("lsb_release") != "") 
    print(paste("Ubuntu:",system("lsb_release -rc", intern=TRUE)[1]))
#print(paste("Rstudio version:", rstudio::versionInfo()$version))  # does not work in Rmd

```
```{r sys-2}

sessionInfo()

```

The RMarkdown file has been published as a [Github Project](https://github.com/ronasta/RepData_PeerAssessment2)  
  
The report was produced using `RStudio/knittr`  
**on `r format(Sys.time(), "%F at %H:%M:%S (%Z, GMT%z)")`**  
